Computational Prediction of Boundaries of 3D Genomic Domains in Class Imbalance Settings

### 2019-02-27 Manuscript/manuscript.Rmd
manuscript edits

The distinct patterns of epigenomic marks, binding peaks of different factors, and actively transcribed genes prompt the use of computational approaches to identify which epigenomic features are most predictive of TAD boundaries. Current methods rely on the use of classification algorithms to draw associations between the spatial relationship of functional genomic elements and regions that include TAD boundaries [@mourad2016computational; @hong2017computational; @huang2015predicting]. Although several useful insights have been made, such as the convergent binding of CTCF [@ong2014ctcf; @ghirlando2016ctcf], many methods ignore key characteristics of domain data, such as the impact of heavily imbalanced number of boundaries vs. other regions. In a classification setting, when one is attempting to discriminate between regions with and without TAD boundaries, the class imbalance issue will contribute to small imbalance ratios (defined as the ratio of minority to majority classes) [@Orriols-Puig2008]. Such heavily imbalanced datasets are known to adversely impact the performance of the classifiers as the learned model is biased towards the majority class to minimize the overall error rate [@lusa2010class;@chen2013influence;@jeni2013facing]. Consequently, careful consideration must be taken when deciding which performance metric(s) to use for evaluation of predictive models, especially in the case of imbalanced data.

@@

In addition to employing suitable metrics for evaluating models validated on imbalanced data, there also exist several algorithmic or data level solutions. Algorithmic level solutions involve modifyng existing classifiers in order to curtail the bias induced by imbalanced classes [@estabrooks2004multiple; @chawla2003smoteboost; @chen2004using; @khoshgoftaar2007empirical]. Data level solutions mostly constitute as resampling techniques, in which random sampling is used to either replicate or remove data points in order to create a more balanced dataset [@chawla2004special; @chawla2009data]. It has been shown that quite simple resampling techniques can drastically improve model performance in the face of imbalanced data [@japkowicz2002class; @dubey2014analysis]. For the purposes of this study, we focus our attention toward data level solutions so that all results may be classifier-independent, and, thus, more easily comparable.

The other aspect of TAD boundary modelling not considered by the current methods is the concept of genomic distance. The typical approaches for modeling the association between TAD boundaries and epigenomic elements include overlap counts or percents [??? Do we have refs for this?; !!!should I repeat past used references here?]. Counts refer to the total number of instances where a genomic annotation overlaps with a genomic region of interest, while percents refers to the percent of overlap that occurred (usually in terms of base pairs). The use of distances from TAD boundaries to regions of various genomic features, measured in the number of bases between them, has largely been ignored. However, these distance characteristics have the potential to offer a more accurate spatial representation of which genomic features are associated with TAD boundaries.

@@

In this study, we have developed an ensemble framework for predicting TAD boundaries in order to systematically compare many data driven solutions used to address the issues presented above. We evaluated Hi-C data for both the GM12878 and K562 cell line obtained at 10 kb, 25 kb, 50 kb, and 100 kb resolution. For each resolution, we examined the inclusion of three different predictor types, including distance as the measure of association between genomic elements. We then tested four different re-sampling techniques aimed at creating balanced classes. Noninformative features were removed using a variable reduction algorithm and random forest classifications were performed to assess model performance, and investigate which genomic annotations were most associated with the TAD boundaries for a given system from the complete ensemble.  Model performance was evaluated using area under the precision-recall curve (AUPRC), F1-Score, Youden's index (YI), and Matthew's correlation coefficient (MCC).

@@

1. **Overlap Counts (OC)**: For each genomic bin, the total number of instances where a bin overlapped with genomic annotation regions (features) was calculated. In the case of no overlaps the value is set as 0.

@@

5. **ROSE** (Random Over-Sampling Examples): Similar to SMOTE, ROSE combines techniques of over-sampling and under-sampling. However, contrary to SMOTE, all of the data generated by ROSE is synthetic. The data is created by sampling from one of the two classes, then generating a new example in its neighborhood using a smoothed bootstrap approach. The shape of the neighborhood is determined by the kernel density estimate of the two classes [@lunardon2013r]. Another important distinction between ROSE and SMOTE is that, for ROSE, the creation of perfectly balanced classes is unlikely due to the randomness in sampling between the two classes. The `ROSE` R package was used to perform ROSE (version 0.0.3).

To determine the best predictor type we performed separate random forest models on each set of predictor types and evaluated the precision-recall curves, as well as the subsequent AUPRC. To confirm the optimality of the feature engineering, this was performed for each combination of re-sampling technique, resolution, and cell line.

@@

Recursive feature elimination (RFE) was then used to determine the optimal number of features for models with the best performing predictor type. For each set of predictor type, random forest models were performed recursively on subsets of the feature space. The specific subsets of features were chosen as powers of 2 for computational efficiency. Model performance for each subset was evaluated using AUPRC. The optimal set of featues to include in downstream analyses was determined to be the subset in which the AUPRC leveled off. Again, to confirm that results were consistent, the reduction process was repeated for each combination of re-sampling technique, resolution, and cell line.

@@

Once the appropriate predictor type and set of features were determined, a final set of random forest classification models were performed to fully compare re-sampling techniques. To reduce bias due to random dataset generation, 10-fold cross validation was used. The default number of features to consider at each node of the random forest algorithm was set as the square root of the number of features in the model. Likewise, the number of ensemble trees to aggregate was set at 500. All random forest algorithms were implemented in R using the `caret` package version 6.0 [@kuhn2012caret]. Once the model was implemented, it was assessed using the test set, and performance metrics were then recorded.

@@

We evaluated and compared multiple random forest classification algorithms built on domain data extracted from Hi-C datasets for 2 cell lines C = {GM12878, K562}, at 4 different resolutions R = {10 kb, 25 kb, 50 kb, 100 kb}. For each classifier, we compared 3 predictor types P = {OC, OP, Distance} and 5 class balancing techniques B = {None, RUS, ROS, SMOTE, ROSE}. A system produced from the ensemble framework was defined as: E = {c, r, p, b}, where $c \in C, \quad R \in R, \quad p \in P,$ and $b \in$ B. Letting |X| denote the cardinality of each set, we evaluated $|C| \times |R| \times |P| \times |B|$ different ensemble systems, totaling X (=$x \times 4 \times 3 \times 5$) separate models. For each model, we aimed to assess both performance and identify the genomic annotations that were most important in correctly labeling the genomic bins that overlapped with flanked TAD boundaries.

We first examined the class imbalance of the different data resolutions (Supplementary Table 5). [??? What table?; !!!fixed] While the total number of genomic bins decreased as resolution increased, the imbalance ratios remained relatively stable. We found that there was an average imbalance ratio of 0.154 (SD=0.022) and 0.137 (SD=0.0181) across resolutions for cell lines GM12878 and K562 respectively. [??? At which resolution?; !!!fixed] The low imbalance ratios indicate a wide disparity between majority and minority classes, and therefore suitable resampling techniques are needed.

## DISTANCE TYPE PREDICTORS OUTPERFORM OTHER PREDICTOR-TYPES

@@

Recursive feature elimination was used to establish the optimal number of distance-type features to include in downstream analyses. It was determined that, for 10 kb resolution data, near optimal performance was obtained for models with only 32 predictors, across each resampling technique and for each cell line (Figures 9-13). Results remained consistent across the other data resolutions (Supplementary Figures 24-38). Therefore, the feature space was reduced for subsequent models to the top 32 predictors, respective to cell line, resolution, and resampling technique.

>>>>

The distinct patterns of epigenomic marks, binding peaks of different factors, and actively transcribed genes prompt the use of computational approaches to identify which epigenomic features are most predictive of TAD boundaries. Current methods rely on the use of classification algorithms to draw associations between the spatial relationship of functional genomic elements and regions that include TAD boundaries. Although several useful insights have been made, such as the convergent binding of CTCF [@ong2014ctcf; @ghirlando2016ctcf], many methods ignore key characteristics of domain data, such as the impact of heavily imbalanced number of boundaries vs. other regions. In a classification setting, when one is attempting to discriminate between regions with and without TAD boundaries, the class imbalance issue will contribute to small imbalance ratios (defined as the ratio of minority to majority classes) [@Orriols-Puig2008]. Such heavily imbalanced datasets are known to adversely impact the performance of the classifiers as the learned model is biased towards the majority class to minimize the overall error rate [@lusa2010class;@chen2013influence;@jeni2013facing]. Consequently, careful consideration must be taken when deciding which performance metric(s) to use for evaluation of predictive models, especially in the case of imbalanced data.

@@

In addition to employing suitable metrics for evaluating models validated on imbalanced data, there also exist several algorithmic or data level solutions. Algorithmic level solutions involve modifyng existing classifiers in order to curtail the bias induced by imbalanced classes [@estabrooks2004multiple; @chawla2003smoteboost; @chen2004using; @khoshgoftaar2007empirical]. Data level solutions, on the other hand, mostly constitute as resampling techniques, in which random sampling is used to either replicate or remove data points in order to create a more balanced dataset [@chawla2004special; @chawla2009data]. It has been shown that quite simple resampling techniques can drastically improve model performance in the face of imbalanced data [@japkowicz2002class; @dubey2014analysis]. For the purposes of this study, we focus our attention toward data level solutions so that all results may be classifier-independent, and, thus, more easily comparable.

Another aspect of TAD boundary modelling not currently being explored are the various feature engineering methods used to build the feature space of models. Different techniques can be employed to measure the relationship between TAD boundaries and genomic annotations. Typical approaches for modeling the association between TAD boundaries and epigenomic elements include overlap counts or percents [@mourad2016computational; @hong2017computational; @huang2015predicting]. Counts refer to the total number of instances where a genomic annotation overlaps with a genomic region of interest, while percents refers to the percent of overlap that occurred (usually in terms of base pairs). The use of distances from TAD boundaries to regions of various genomic features, measured in the number of bases between them, has largely been ignored. However, these distance characteristics have the potential to offer a more accurate spatial representation of which genomic features are associated with TAD boundaries. Likewise, any direct comparisons of the different feature engineering techniques remain unresearched.

@@

In this study, we have developed an ensemble framework for predicting TAD boundaries in order to systematically compare many data driven solutions used to address the issues presented above. We evaluated Hi-C data for both the GM12878 and K562 cell line obtained at 10 kb, 25 kb, 50 kb, and 100 kb resolution. For each resolution, we examined the inclusion of three different predictor types, including distance as the measure of association between genomic elements. We then tested four different re-sampling techniques aimed at creating balanced classes. Noninformative features were removed using a feature reduction algorithm and random forest classifications were performed to assess model performance, and investigate which genomic annotations were most associated with the TAD boundaries for a given system from the complete ensemble.  Model performance was evaluated using area under the precision-recall curve (AUPRC), F1-Score, Youden's index (YI), and Matthew's correlation coefficient (MCC).

@@

1. **Overlap Counts (OC)**: For each genomic bin, the total number of instances (bases) where a bin overlapped with genomic annotation regions (features) was calculated. In the case of no overlaps the value is set as 0.

@@

<!--
5. **ROSE** (Random Over-Sampling Examples): Similar to SMOTE, ROSE combines techniques of over-sampling and under-sampling. However, contrary to SMOTE, all of the data generated by ROSE is synthetic. The data is created by sampling from one of the two classes, then generating a new example in its neighborhood using a smoothed bootstrap approach. The shape of the neighborhood is determined by the kernel density estimate of the two classes [@lunardon2013r]. Another important distinction between ROSE and SMOTE is that, for ROSE, the creation of perfectly balanced classes is unlikely due to the randomness in sampling between the two classes. The `ROSE` R package was used to perform ROSE (version 0.0.3).
-->

@@

A regularization technique using a combination of the $l_{1}$ and $l_{2}$ penalties, known as the elastic-net, was used to reduce the feature space. The elastic-net was chosen in order to suitably reduce the feature space while also maintaining the inclusion of multiple correlated predictors. For each elastic-net regularization, 10-fold cross-validation was used to tune the parameters that controlled the penalty terms. All elastic-net algorithms were implemented in R using the `caret` package version 6.0 [@kuhn2012caret].

<!--
Recursive feature elimination (RFE) was then used to determine the optimal number of features for models with the best performing predictor type. For each set of predictor type, random forest models were performed recursively on subsets of the feature space. The specific subsets of features were chosen as powers of 2 for computational efficiency. Model performance for each subset was evaluated using AUPRC. The optimal set of featues to include in downstream analyses was determined to be the subset in which the AUPRC leveled off. Again, to confirm that results were consistent, the reduction process was repeated for each combination of re-sampling technique, resolution, and cell line. 
-->

@@

Once the appropriate predictor type and set of features were determined, a final set of random forest classification models were performed to fully compare re-sampling techniques. To reduce bias due to random dataset generation, 10-fold cross validation was used <!--to tune the number of features to consider at each node and the number of ensemble trees to aggregate-->.The default number of features to consider at each node of the random forest algorithm was set as the square root of the number of features in the model. Likewise, the number of ensemble trees to aggregate was set at 500. All random forest algorithms were implemented in R using the `caret` package version 6.0 [@kuhn2012caret]. Once the model was implemented, it was assessed using the test set, and performance metrics were then recorded.

@@

We evaluated and compared multiple random forest classification algorithms built on domain data extracted from Hi-C datasets for 2 cell lines C = {GM12878, K562}, at 4 different resolutions R = {10 kb, 25 kb, 50 kb, 100 kb}. For each classifier, we compared 3 predictor types P = {OC, OP, Distance} and 5 class balancing techniques B = {None, RUS, ROS, SMOTE, ROSE}. A system produced from the ensemble framework was defined as: E = {c, r, p, b}, where $c \in C, \quad R \in R, \quad p \in P,$ and $b \in$ B. Letting |X| denote the cardinality of each set, we evaluated $|C| \times |R| \times |P| \times |B|$ different ensemble systems, totaling X (=$x \times 4 \times 3 \times 5$) separate models. For each model, we aimed to assess both model performance and identify the genomic annotations that were most important in correctly labeling the genomic bins that overlapped with flanked TAD boundaries.

## IMBALANCE RATIOS REMAINED SEVERE AND CONSISTENT ACROSS RESOLUTIONS

We first examined the class imbalance of the different data resolutions for each cell line (Supplementary Table 5). While the total number of genomic bins decreased as resolution increased, the imbalance ratios remained relatively stable. We found that there was an average imbalance ratio of 0.154 (SD=0.022) and 0.137 (SD=0.0181) across resolutions for the GM12878 and K562 cell lines respectively. The low imbalance ratios indicate a wide disparity between majority and minority classes, and therefore suitable resampling techniques were needed.

## MANY GENOMIC ANNOTATIONS WERE HIGHLY CORRELATED

When assessing the collinearity of the genomic annotations, it was found that the total number of bases that the number overlapped bases between genomic bins flanked by a TAD boundary were highly correlated between many genomic annotations (). That is, many genomic annotations tended to colocalize near eachother at the boarders of domains. This was present across both resolution and cell line. Specifically, we found that the CTCF, Rad21, Smc3ab, and Znf143 transcription factor binding sites were highly colocalized near TAD boundaries. Likewise, many histone modifications also tended to appear near TAD boundaries in conjunction with one another including H3K4me1, H3K4me2, H3K4me3, H3K9ac, and H3K27ac. Furthermore, particular chromatin segmentation states, inclduing heterochromatin, repressed chromatin, poised promotors, and repetitive copy number variations, were found to be highly correlated with on another, yet inversely correlated with many transcription factor binding sites. The high correlations among many of the genomic annotations gives precedent for the use of an elastic-net regularization.

<!--
## DISTANCE TYPE PREDICTORS OUTPERFORM OTHER PREDICTOR-TYPES

@@

Recursive feature elimination was used to establish the optimal number of distance-type features to include in downstream analyses. It was determined that, for 10 kb resolution data, near optimal performance was obtained for models with only 32 predictors, across each resampling technique and for each cell line (Figures 9-13). Results remained consistent across the other data resolutions (Supplementary Figures 24-38). Therefore, the feature space was reduced for subsequent models to the top 32 predictors, respective to cell line, resolution, and resampling technique. 
-->
